---
title: "Survival Analysis"
author: "Robin Manz, Matias Tolppanen"
date: "2024-11-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(ggplot2)
library(survival)
library(splines)
```

## R Markdown

## Renal transplantation and risk of graft failure

outcome:
gstatus         = graft-rejection (=1) or not (=0)
gsurv           = number of months the patient was followed until rejection or censoring
pstatus         = patient-status at end of follow-up (1=died, 0=alive)
psurv           = number of months the patient was followed until death or censoring

covariates/risk factors:
acclft          = age of the patient at transplantation (years)
dgf             = delayed graft function (dgf or 0=no dgf)
aantalre        = number of acute rejection-treatments during the first 6 months post-transplantation (0, 1, 2 or 3)
creat           = serum creatinine at six-weeks post-transplantation (mumol/L) (marker of kidney-function)
predias         = diastolic bloodpressure pre-transplantation
prac            = panel reactive antibodies (marker of inflammation)
uprotein        = protein in urine at six-weeks post-transplantation  (geen,  sp/+,  >+) (marker of kidney function)
cregsh          = number of cross-reactive groups (public HLA epitopes) of the major histocompatibility complex class I shared between donor and recipient

c("aantalre" , "acclft" , "dgf" ,"creat", "predias", "prac", "uprotein", "cregsh")]
```{r}
# Loading data
# Load dataset
load(file="renaltx.Rdata")

# show the data available, the data we are looking for is in d
ls()

# remove missing values
d <- d %>% drop_na()
```

## Kaplan Meier curve

```{r} 
# user the Surv() mehtod to extract the sourvival proportion
Y <- Surv(time = d$gsurv, event = d$gstatus == 1)
summary(Y)

# plot the Surv time
plot(Y, lty = c("solid", "dashed", "dashed"), col = c("black", "grey", "grey"), xlab = "Survival Time In Days", ylab = "Graft rejection probabilities for all patients", main="Kaplan-Meier Curve")
```
## Univariate Analyiss: Check for linearity using splines
If min(AIC(ns(df>1)))< (AIC(ns(df=1))) linearity is violated and will be fixed by using ns with df == x

```{r}

plot_splines <- function(covariate) {
  print(paste("############", covariate, "##########"))
  psplinespvals <- c()
  aic=c()
  i <- 1
  # acclft
  # test out 1 to 12 degrees of freedom
  for (j in 1:10) {
    # compute the AIC of the cox-model with i defrees of freedom for the approach
    formula <- as.formula(paste("Surv(gsurv, gstatus) ~ ns(", covariate, ", df=", toString(j), ")"))
    aic[j] = AIC(coxph(formula, data=d))
  }
  
  # plot the aics for the splines, the lowest AIC is the best fit
  plot(aic, type="b", main=covariate, xlab="natural spline penalty (df)")
  # get the best pspline value 
  formula <- as.formula(paste("Surv(gsurv, gstatus) ~ pspline(", covariate, ", data=d)"))
  qq=summary(coxph(formula, data=d))
  psplinespvals[i] = qq$coefficients[2,6] # ? 
  # add a text with the positions at x and y (you can decipher that on your own (: )
  text(x=1, y=min(aic,na.rm=T)+(max(aic,na.rm=T)-min(aic,na.rm=T))/2, 
  paste("pspline nonlin p = ", round(psplinespvals[i],4)), adj=0, col=4, cex=0.9)
  
  # get the minimum AIC
  best_df = which.min(aic)
  # Test linearity by comparing AIcs
  # Linear model
  formula <- as.formula(paste("Surv(gsurv, gstatus) ~ ", covariate))
  linear_model <- coxph(formula, data = d)
  aic_linear <- AIC(linear_model)
  
  # Spline model with 3 degrees of freedom (adjust as needed)
  formula <- as.formula(paste("Surv(gsurv, gstatus) ~ ns(", covariate, ", df= ", toString(best_df), ")"))
  spline_model <- coxph(formula, data = d)
  aic_spline <- AIC(spline_model)
  
  # Compare AICs
  print(paste("AIC (Linear):", round(aic_linear, 2)))
  print(paste("AIC (Spline):", round(aic_spline, 2)))
  
  # Likelihood Ratio Test
  lrt <- anova(linear_model, spline_model, test = "Chisq")
  print(lrt)
  
  # Visual inspection
  # plot(spline_model)
  
  # check for proportionality
  cox_l=cox.zph(linear_model)
  plot(cox_l)
  # print(summary(linear_model))
  print("cox_zph linear")
  print(cox_l)
  
  cox_s=cox.zph(spline_model)
  plot(cox_s)
  # print(summary(spline_model))
  print("compare the p-value for proportionality")
  print("cox_zph linear")
  print(cox_l)
  print("cox_zph splines")
  print(cox_s)
}
covs_continu <- c("aantalre" , "acclft" ,"creat", "predias", "prac", "cregsh")

for(x in covs_continu) {
  plot_splines(x)
}
```

Result: acclft violates the linearity and therefore we use ns(acclft, df=4) based on the charts, this result is significant when comparing it to a linear model

From this point on we assume all variables are linearly related and we use ns(acclft, df=4) for acclft as a fix. 


Proportionality is violated for acclft (p=0.00014) and aantalre (0.0036)
## Fix proportionality for acclft and aantalre with landmarking
We apply landmarking to the time-axis for our model (gsurv)

```{r}
library(survival)

perform_landmarking <- function(covariate) {
  formula <- as.formula(paste("Surv(gsurv, gstatus) ~", covariate))
  # compute landmarks (we chose 5 intervals
  landmarks <- seq(min(d$gsurv), max(d$gsurv), length.out=3)
  #landmarks <- c(0, 30, 60, 90, 120, 150, 180)
  print(landmarks)
  # do the survival time split
  dx = survSplit(formula, d, cut=landmarks, episode ="timegrp", start="tstart", end="tstop")
  # exact the same results as the analysis with the original dataset in d1
  View(dx)
  print("#### Normal model ######")
  cox_n <- coxph(Surv(gsurv, gstatus) ~ ns(acclft, 4), data = d)
  summary(cox_n)
  print(AIC(cox_n))
  cox_n_szph=cox.zph(cox_n)
  print(cox_n_szph)
  
  print("###### ladmark model
    ########")
  if (covariate == "acclft") {
    formula <- as.formula("Surv(tstart, tstop, gstatus) ~ ns(acclft, 4)")
  } else {
    formula <- as.formula(paste("Surv(tstart, tstop, gstatus) ~", covariate))
  }
  cox_land <- coxph(formula, data= dx)
  summary(cox_land)
  print(cox_land)
  cox_land_szph=cox.zph(cox_land)
  print(cox_land_szph)
  
  print("######## landmarkmodel with time-intervals #######")
  # let the effect of creatinine vary with the time-intervals
  if (covariate == "acclft") {
    f <- as.formula("Surv(tstart, tstop, gstatus) ~ ns(acclft, 4) * strata(timegrp) ")
  } else {
    f <- as.formula(paste("Surv(tstart, tstop, gstatus) ~", covariate, "* strata(timegrp) "))
  }

  int_cox <- coxph(f , data= dx)
  print(summary(int_cox))
  print(AIC(int_cox))
  # TODO fix too small number bug
  cox_int_szph=cox.zph(int_cox)
  print(cox_int_szph)
  
  validate(int_cox, method="boot", B=100) 

}

c <- c("aantalre", "acclft")
for (x in c) {
  perform_landmarking(x)
}

# effect of creatinine varies with time, effect of sex not ..
#summary(coxph(Surv(tstart, tstop, gstaus) ~ geslacht + 
      #creatinine * strata(timegrp), data= d1x))


```
## Multivariate Regression 

```{r}
plot_slines_multi <- function (cov) {
  coxes <- coxph(Surv(gsurv,gstatus) ~ aantalre + acclft + dgf + creat + predias + prac + uprotein + cregsh, data = d)
  
  coxesa=cox.zph(coxes)
  par(mfrow=c(2,2))
  plot(coxesa[1])
  plot(coxesa[2])
  plot(coxesa[3])
  plot(coxesa[4])
  plot(coxesa[5])
  plot(coxesa[6])
  plot(coxesa[7])
  plot(coxesa[8])
  par(mfrow=c(1,1))
  coxesa
}


covs_continu <- c("aantalre" , "acclft" ,"creat", "predias", "prac", "cregsh")

for (x in covs_continu) {
  
}

```



