---
title: "MAM03_Week2"
author: "Robin Manz, Matias Tolppanen"
date: "2024-11-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("tidyr")
library(tidyr)
```

## Renal transplantation and risk of graft failure

outcome:

            gstatus             = graft-rejection (=1) or not (=0)

            gsurv                = number of months the patient was followed until rejection or censoring

            pstatus             = patient-status at end of follow-up (1=died, 0=alive)

            psurv                = number of months the patient was followed until death or censoring

 

covariates/risk factors:

            acclft              = age of the patient at transplantation (years)

            dgf                   = delayed graft function (dgf or 0=no dgf)

            aantalre           = number of acute rejection-treatments during the first 6 months post-transplantation (0, 1, 2 or 3)

            creat                = serum creatinine at six-weeks post-transplantation (mumol/L) (marker of kidney-function)

            predias            = diastolic bloodpressure pre-transplantation

            prac                 = panel reactive antibodies (marker of inflammation)

            uprotein           = protein in urine at six-weeks post-transplantation  (geen,  sp/+,  >+) (marker of kidney function)

            cregsh              = number of cross-reactive groups (public HLA epitopes) of the major histocompatibility complex class I shared between donor

                                        and recipient

```{r}
# Load dataset
load(file="renaltx.Rdata")

# show the data available
ls()
# View(d)

```

## Descriptives

```{r pressure, echo=FALSE}
d_rejected <- subset(d, gstatus == 1)
d_normal <- subset(d, gstatus == 0)

# histgrams for the distrubutions of all the variables 
# outcomes
hist(d$gsurv, breaks = 20, main="months till rejection or censoring (rejected)")
hist(d$psurv, breaks=20, main="months till death or censoring")
hist(d$pstatus, breaks=2, main="Patient status 1 = dead, 0 = alive")
hist(d$gstatus, breaks=2, main="Graft rejection status 1 = rejected, 0 = not rejected")

# predictors
# aantalre + acclft + dgf + creat + predias + prac + uprotein + cregsh
hist(d$acclft, breaks=20, main="Age of transplantation years (acclft)")
# hist(d$dgf, breaks=20, main="Delay graft function (dgf)")
hist(d$aantalre, breaks=4, main="Number of rejection treatments within \n the first 6 months(aantalre)")
hist(d$creat, breaks=20, main="Kidney function (serum creatinine mumol/L)")
hist(d$predias, breaks=20, main="Diastolic blood pressure pre transplantation")
hist(d$prac, breaks=20, main="Inflamation markers (Prac)")
# hist(d$uprotein, breaks=20, main="Protein in urine")
hist(d$cregsh, breaks=8, main="Number of cross-reactive groups")


```
1. Predictor variables acclft approx. normal distr.
2. aantalre decreased over time, skewed to the right
3. createnine skewed to the left
4. blood pressure normal distributed
5. prac skewed to the right
6. cross-reactive groups normal 

## Further analysis
``` {r}
# remove missing values
print(paste("Missing values", toString(sum(is.na(d)))))
d <- d %>% drop_na()
print(paste("Missing values", toString(sum(is.na(d)))))

print(paste("Number of patients alive", nrow(subset(d, pstatus==1))))
print(paste("Number of patients dead", nrow(subset(d, pstatus==0))))
print(paste("Number of patients with graft rejection", nrow(subset(d, gstatus==1))))
print(paste("Number of patients with no graft rejection", nrow(subset(d, gstatus==0))))

# confusion table
table(d$gstatus, d$pstatus)
# do not know if this is right i am trying to plot the incidence for any given month

get_incidence_graph_rejection <- function(timepoints) {
  # save the incidences
  incidences <- c()
  # for each timepoint
  for (i in timepoints) {
    
    n_ev_in_period <- nrow(subset(d, gstatus==1 & gsurv < i))
    # total number
    n_patients_at_risk <- nrow(d)
    incidences[i] <- n_ev_in_period/ n_patients_at_risk
  }
  return(incidences)
}

inci <- get_incidence_graph_rejection(seq(1, max(d$gsurv)))

plot(inci, main="Incidences per month")

```
## Kaplan Meier Curve

``` {r} 
library(ggplot2)
library(survival)

# user the Surv() mehtod to extract the sourvival proportion
Y <- Surv(time = d$gsurv, event = d$gstatus == 1)
summary(Y)

# Splitting it up between predictors (not used)
# kmfit = survfit(Y ~ addicts$clinic)

# plot the Surv time
plot(Y, lty = c("solid", "dashed", "dashed"), col = c("black", "grey", "grey"), xlab = "Survival Time In Days", ylab = "Graft rejection probabilities for all patients")

?survfit
# plot graft rejection 
d_g_rejected <- subset(d, gstatus == 1)
y2 <- Surv(time=d_g_rejected$gsurv, event = d_g_rejected$gstatus)

plot(y2, lty = c("solid", "dashed", "dashed"), col = c("black", "grey", "grey"), xlab = "Survival Time In Days", ylab = "Graft rejection probabilities for people with event")

```

## checks for linearity
```{r}
plot(d$acclft, d$gsurv)
model1 = lm(gsurv~acclft, data=d)
lines(d$acclft,predict(model1),col=2,lwd=2)
hist(model1$residuals)
plot(model1)

plot(d$aantalre, d$gsurv)
model2 = lm(gsurv~aantalre, data=d)
lines(d$aantalre,predict(model2),col=2,lwd=2)
hist(model2$residuals)
plot(model2)

plot(d$creat, d$gsurv)
model3 = lm(gsurv~creat, data=d)
lines(d$creat,predict(model3),col=2,lwd=2)
hist(model3$residuals)
plot(model3)

plot(d$predias, d$gsurv)
model4 = lm(gsurv~predias, data=d)
lines(d$predias,predict(model4),col=2,lwd=2)
hist(model4$residuals)
plot(model4)#

plot(d$prac, d$gsurv)
model5 = lm(gsurv~prac, data=d)
lines(d$prac,predict(model5),col=2,lwd=2)
hist(model5$residuals)
plot(model5)

plot(d$cregsh, d$gsurv)
model6 = lm(gsurv~cregsh, data=d)
lines(d$cregsh,predict(model6),col=2,lwd=2)
hist(model6$residuals)
plot(model6)

# aantalre + acclft + dgf + creat + predias + prac + uprotein + cregsh
# might all be just bollucks

```
A: all decently linear ?

## List of the regression models
| name  | assumption | suitable (yes/no) |
| --- | --- | --- |
| Cox proportional hazards model | linearity of continuous predictors, proportionality, indipendeance of observations | yes ? |
| Fine-Gray subdistribution hazard model | TODO | TODO | 

## check for proportionality 
```{r}

coxes <- coxph(Surv(gsurv,gstatus) ~ aantalre + acclft + dgf + creat + predias + prac + uprotein + cregsh, data = d)

coxesa=cox.zph(coxes)
par(mfrow=c(2,2))
plot(coxesa[1])
plot(coxesa[2])
plot(coxesa[3])
plot(coxesa[4])
plot(coxesa[5])
plot(coxesa[6])
plot(coxesa[7])
plot(coxesa[8])
par(mfrow=c(1,1))
coxesa
```
proportionality results:
method,     chisqr, df, p
aantalre  7.8944  1 0.00496
acclft   14.0144  1 0.00018
have p-values < 0.05 and indicate violation of the proportionality
both also have a non-parallel plot
The others are approximaltely parallel


## Trying to fix proportionality for univariate analysis

```{r} 


covs <- d[,c("aantalre" , "acclft" , "dgf" ,"creat", "predias", "prac", "uprotein", "cregsh")]
library(splines)
psplinespvals <- c()
aic=c()
i <- 1
# acclft
# test out 1 to 12 degrees of freedom
for (j in 1:12) {
  # compute the AIC of the cox-model with i defrees of freedom for the approach
  aic[j] = AIC(coxph(Surv(gsurv, gstatus) ~ ns(acclft, df=j), data=d))
}

# plot the aics for the splines, the lowest AIC is the best fit
plot(aic, type="b", main="acclft", xlab="natural spline penalty (df)")
# get the best pspline value 
qq=summary(coxph(Surv(gsurv, gstatus) ~ pspline(acclft), data=d))
psplinespvals[i] = qq$coefficients[2,6] # ? 
# add a text with the positions at x and y (you can decipher that on your own (: ))
text(x=1, y=min(aic,na.rm=T)+(max(aic,na.rm=T)-min(aic,na.rm=T))/2, 
paste("pspline nonlin p = ", round(psplinespvals[i],4)), adj=0, col=4, cex=0.9)

# creat
for (j in 1:12) {
  aic[j] = AIC(coxph(Surv(gsurv, gstatus) ~ ns(creat, df=j), data=d))}
  plot(aic, type="b", main="creat", xlab="natural spline penalty (df)")
  qq=summary(coxph(Surv(gsurv, gstatus) ~ pspline(creat), data=d))
  psplinespvals[i] = qq$coefficients[2,6]
  text(x=1, y=min(aic,na.rm=T)+(max(aic,na.rm=T)-min(aic,na.rm=T))/2, 
             paste("pspline nonlin p = ", round(psplinespvals[i],4)), adj=0, col=4, cex=0.9)
  
# predias
for (j in 1:12) {
  aic[j] = AIC(coxph(Surv(gsurv, gstatus) ~ ns(predias, df=j), data=d))}
  plot(aic, type="b", main="predias", xlab="natural spline penalty (df)")
  qq=summary(coxph(Surv(gsurv, gstatus) ~ pspline(predias), data=d))
  psplinespvals[i] = qq$coefficients[2,6]
  text(x=1, y=min(aic,na.rm=T)+(max(aic,na.rm=T)-min(aic,na.rm=T))/2, 
             paste("pspline nonlin p = ", round(psplinespvals[i],4)), adj=0, col=4, cex=0.9)

# prac
for (j in 1:12) {
  aic[j] = AIC(coxph(Surv(gsurv, gstatus) ~ ns(prac, df=j), data=d))}
  plot(aic, type="b", main="prac", xlab="natural spline penalty (df)")
  qq=summary(coxph(Surv(gsurv, gstatus) ~ pspline(prac), data=d))
  psplinespvals[i] = qq$coefficients[2,6]
  text(x=1, y=min(aic,na.rm=T)+(max(aic,na.rm=T)-min(aic,na.rm=T))/2, 
             paste("pspline nonlin p = ", round(psplinespvals[i],4)), adj=0, col=4, cex=0.9)
  
  
# cregsh
for (j in 1:12) {
  aic[j] = AIC(coxph(Surv(gsurv, gstatus) ~ ns(cregsh), data=d))}
  plot(aic, type="b", main="cregsh", xlab="natural spline penalty (df)")
  qq=summary(coxph(Surv(gsurv, gstatus) ~ pspline(cregsh), data=d))
  psplinespvals[i] = qq$coefficients[2,6]
  text(x=1, y=min(aic,na.rm=T)+(max(aic,na.rm=T)-min(aic,na.rm=T))/2, 
             paste("pspline nonlin p = ", round(psplinespvals[i],4)), adj=0, col=4, cex=0.9)
  
  
```
## acclft isnt linear and need splines with df = 4

```{r}
model2=coxph(Surv(gsurv, gstatus) ~ pspline(acclft), data=d)
summary(model2)

### ???
cox_acclft <- coxph(Surv(gsurv, gstatus) ~ ns(acclft,4), data = d)
cox_s=cox.zph(cox_acclft)
plot(cox_s)
summary(cox_acclft)

```
concodance psplines 0.556
concodance natural splines 0.563

## Univariate Analysis
```{r}

# Define the list of predictors
predictors <- c("aantalre", "acclft", "dgf", "creat", "predias", "prac", "uprotein", "cregsh")

# Initialize a data frame to store results
results <- data.frame(
  Predictor = character(),
  HR = numeric(),
  `Lower CI` = numeric(),
  `Upper CI` = numeric(),
  p.value = numeric(),
  stringsAsFactors = FALSE
)

# Fit Cox model for each predictor
for (pred in predictors) {
  # Define the formula for the univariable model
  if (pred == "acclft") {
    # Use natural splines for `acclft`
    formula <- as.formula("Surv(gsurv, gstatus) ~ ns(acclft, 4)")
  } else {
    # Standard formula for other predictors
    formula <- as.formula(paste("Surv(gsurv, gstatus) ~", pred))
  }
  
  # Fit the Cox proportional hazards model
  cox_model <- coxph(formula, data = d)
  
  # Extract hazard ratio, confidence interval, and p-value
  summary_cox <- summary(cox_model)
  hr <- summary_cox$coef[1, "exp(coef)"]
  lower_ci <- summary_cox$conf.int[1, "lower .95"]
  upper_ci <- summary_cox$conf.int[1, "upper .95"]
  p_value <- summary_cox$coef[1, "Pr(>|z|)"]
  
  # Store the results
  results <- rbind(results, data.frame(
    Predictor = pred,
    HR = hr,
    `Lower CI` = lower_ci,
    `Upper CI` = upper_ci,
    p.value = round(p_value, 6)
  ))
}

# Display the results
print(results)

```
## what kind of relationship ? 

```{r}
# plot the predictions
cox_acclft_model <- coxph(Surv(gsurv, gstatus) ~ ns(acclft,4), data = d)
preds <- predict(cox_acclft_model, type="lp")
plot(d$acclft, preds)
text(x = max(d$acclft) * 0.7, y = max(preds) * 0.8, labels=paste("AIC =", round(AIC(cox_acclft_model), 2), col = "blue"))
summary(cox_acclft_model)

cox_acclft_log_model <- coxph(Surv(gsurv, gstatus) ~ log(acclft), data = d)
preds <- predict(cox_acclft_log_model, type="lp")
plot(d$acclft, preds)
text(x = max(d$acclft) * 0.7, y = max(preds) * 0.8, labels=paste("AIC =", round(AIC(cox_acclft_log_model), 2), col = "blue"))
summary(cox_acclft_log_model)


# concodance psplines 0.556
# concodance natural splines 0.563
# concordance log model 0.532


qq1=coxph(Surv(gsurv, gstat) ~ acclft, 
        data=d)
cox.zph(qq1)
plot(cox.zph(qq1))
abline(h=0, col=2, lty=3, lwd=3)

```
## Power plot
```{r}
# Initialize AIC container
aicc <- c()

# Define power sequence (λ)
power <- seq(-2, 2, 0.1)

# Loop through each power value and fit the Cox model
for (j in 1:length(power)) {
  # Apply the power transformation to acclft (or any other predictor you're interested in)
  transformed_pred <- d$acclft^power[j]  # Adjust "d$acclft" to your column name
  
  # Fit the Cox model with the transformed predictor
  cox_model <- coxph(Surv(gsurv, gstatus) ~ transformed_pred, data = d)
  
  # Store AIC of the fitted model
  aicc[j] <- AIC(cox_model)
}

# Plot the AIC values against the power values (λ)
plot(power[power != 0], aicc[power != 0], xlab = "lambda", ylab = "AIC", type = "b")

# Find the power that gives the minimum AIC
best_power <- power[which.min(aicc)]
best_power

```

## results Univariate Analysis: 
aantalre(p<0.0001), acclft(p<0.004), creat(p<0,004), uprotein(p<0.002) and cregsh (p<0.001)
are significant






## Cox regression

Assumtions:
1. linearity (can be dropped by using other mehtods)
2. proportionality

``` {r}
library(splines)
#plot(survfit(Surv(gsurv, gstatus) ~ aantalre + acclft + dgf + creat + predias + prac + uprotein + cregsh, data=d))
plot(survfit(Surv(gsurv, gstatus) ~ acclft, data=d))

# perform cox regression assuming linearity pspline for acclft
coxes <- coxph(Surv(gsurv,gstatus) ~ aantalre + ns(acclft, 4) + dgf + creat + predias + prac + uprotein + cregsh, data = d)
?coxph

coxesa=cox.zph(coxes)
par(mfrow=c(2,2))
plot(coxesa[1])
plot(coxesa[2])
plot(coxesa[3])
plot(coxesa[4])
plot(coxesa[5])
plot(coxesa[6])
plot(coxesa[7])
plot(coxesa[8])
par(mfrow=c(1,1))
coxesa

# pspline(acclft) 13.465  4.05 0.0096
?pspline
c <- coxph(Surv(gsurv,gstatus) ~ acclft, data = d)
cs <- cox.zph(c)
plot(cs[1])
cs
c <- coxph(Surv(gsurv,gstatus) ~ pspline(acclft), data = d)
cs <- cox.zph(c)
plot(cs[1])
cs
# results form the cox regression
# aantalre  7.8944  1 0.00496
# acclft   14.0144  1 0.00018
# The lines are not straight so we need to adjust them


# TODO check for linearity 
# plot residuals of modern regression against the mean
```

## Varibale adjustment


Plot age against hazard
```{r}
library(ggplot2)
library(ggsurvfit)
library(survminer)
# Create age groups based on quantiles
breaks <- quantile(d$acclft, probs = c(0,0.1, 0.2,0.3, 0.4,0.5, 0.6,0.7, 0.8,0.9, 1))
d$age_group <- cut(d$acclft, breaks = breaks, include.lowest = TRUE)

surv_fit_age <- survfit(Surv(gsurv, gstatus) ~ pspline(acclft), data = d)

ggsurvplot(surv_fit_age, xlab = "Time", ylab = "Survival Probability", 
           title = "Kaplan-Meier Survival Curves by Age Group", 
           legend.title = "Age Group")

ggsurvplot(surv_fit_age, fun = "cumhaz", xlab = "Time", ylab = "Cumulative Hazard", 
           title = "Cumulative Hazard by Age Group", legend.title = "Age Group")


# time dependeant interaction with acclft(age)
d$age_time <- d$acclft
c <- coxph(Surv(gsurv,gstatus) ~ age_time, data = d)
cs <- cox.zph(c)
plot(cs[1])
```

### Landmarking plot the unnecesairy way

```{r}
# Landmarks max(d$gsurv) == 169.06, min(6.15)
# (max-min) / 5 = 32.58 steps -> intervals of 30
# set landmarks
apply_landmarking <- function(covariate) {
  landmarks=c(0,30,60,90,120,150,180)
  beta=matrix(NA,nrow=(length(landmarks)-1),ncol=2)
  cstat=c()
  for (i in 1:(length(landmarks)-1)) {
    # get subset for the landmark
    dx=subset(d, gsurv > landmarks[i])
    # set gstatus of every partition where the event didnt happen in to 0
    dx$gstatus[dx$gsurv>landmarks[(i+1)]]=0
    # insert new partition ?
    dx$gsurv[dx$gsurv>landmarks[(i+1)]]=landmarks[(i+1)]
    
    form <- as.formula(paste("Surv(gsurv, gstatus)~", covariate))
    help6=summary(coxph(form, data=dx))
    beta[i,1:2]=help6$coefficients[c(1,3)]
    cstat[i]=help6$concordance[1]
  }
  plot(landmarks[1:(length(landmarks)-1)],
        beta[,1], xlab="landmark gsurv (months)",
        ylab="log Hazard Ratio",
        ylim=c(-0.05,0.05))
  abline(h=0,lty=2)
  for (i in 1:nrow(beta)) {
     lines(c(landmarks[i],landmarks[i]),
        c(beta[i,1]-1.96*beta[i,2],
          beta[i,1]+1.96*beta[i,2]))
  }
  for (i in cstat){
    print(cstat)
  }
}

covs <- c("acclft", "aantalre")
for (x in covs) {
  apply_landmarking(x)
}
```

## sophies model

```{r}
landmarks <- seq(min(d$gsurv), max(d$gsurv), length.out=6)

dat_split <- survSplit(Surv(gsurv, gstatus) ~ acclft, d, cut=landmarks, episode = "timegrp", start="tstart", end = "tstop")

split_model <- coxph(Surv(tstart, tstop, gstatus) ~ acclft * strata(timegrp), data=dat_split)

summary(coxph(Surv(gsurv, gstatus) ~ acclft, data=d))

summary(split_model)

AIC(coxph(Surv(gsurv, gstatus) ~ acclft, data=d))

# so we know that the optimal AIC is generated with having the effect of age vary with the age groups
AIC(split_model)

cox.zph(split_model)
```