```{r}
library(survival)
library(survminer)
library(ggplot2)
library(dplyr)
library(splines)
library(rms)
library(cmprsk)
library(e1071)
```

```{r}
# load("/Users/matiastolppanen/Downloads/renaltx.Rdata")
load(file="renaltx.Rdata")
ls()
```
```{r}
# Remove missing data
vars_of_interest <- c("gstatus", "gsurv", "pstatus", "psurv", "acclft", "dgf",
                      "aantalre", "creat", "predias", "prac", "uprotein", "cregsh")
renaltx_clean <- d %>% select(all_of(vars_of_interest)) %>% na.omit()

nrow(renaltx_clean)

renaltx_clean$dgf <- factor(renaltx_clean$dgf)
renaltx_clean$aantalre <- factor(renaltx_clean$aantalre)
renaltx_clean$uprotein <- factor(renaltx_clean$uprotein)
dummies <- model.matrix(~ uprotein - 1, data = renaltx_clean)
renaltx_clean <- cbind(renaltx_clean, dummies)

renaltx_clean
```
Exploratory data analysis
```{r}
# List of quantitative variables and their bin widths
quant_vars <- c("acclft", "creat", "predias", "prac", "cregsh")
binwidths <- list("acclft" = 10, "creat" = 5, "predias" = 10, "prac" = 10, "cregsh" = 1)

# Loop through each variable to calculate skewness, determine transformation, and plot
for (var in quant_vars) {
  # Calculate skewness score
  skewness_score <- skewness(renaltx_clean[[var]], na.rm = TRUE)
  
  # Interpret skewness score
  if (abs(skewness_score) >= 1) {
    cat(var, "is highly skewed (skewness:", skewness_score, "), applying log transformation and replacing original.\n")
    
    # Apply log transformation and replace the original column
    renaltx_clean[[var]] <- log(renaltx_clean[[var]] + 1)
    
    # Plot the log-transformed histogram
    print(ggplot(renaltx_clean, aes_string(x = var)) +
      geom_histogram(fill = "green", color = "black") +
      theme_minimal() +
      labs(title = paste("Log-Transformed Histogram of", var)))
    
  } else {
    cat(var, "has acceptable skewness (skewness:", skewness_score, "), no transformation applied.\n")
    
    # Plot original histogram only if not highly skewed
    print(ggplot(renaltx_clean, aes_string(x = var)) +
      geom_histogram(binwidth = binwidths[[var]], fill = "blue", color = "black") +
      theme_minimal() +
      labs(title = paste("Histogram of", var)))
  }
}
```
Here we did a log transform to creat and prac

Next lets assess linearity using Martingale residuals and splines.
```{r}
# Define the survival object
surv_object <- Surv(time = renaltx_clean$gsurv, event = renaltx_clean$gstatus)

# null Cox model
null_cox_model <- coxph(surv_object ~ 1, data = renaltx_clean)

# martingale residuals
renaltx_clean$martingale_resid <- residuals(null_cox_model, type = "martingale")

# martingale residuals against quantitative variables
for (var in quant_vars) {
  print(ggplot(renaltx_clean, aes_string(x = var, y = "martingale_resid")) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "loess") +
    theme_minimal() +
    labs(x = var, y = "Martingale Residuals",
         title = paste("Martingale Residuals vs", var)))
}
```

```{r}
plot_splines <- function(covariate) {
  print(paste("############", covariate, "##########"))
  psplinespvals <- c()
  aic=c()
  i <- 1
  # acclft
  # test out 1 to 12 degrees of freedom
  for (j in 1:10) {
    # compute the AIC of the cox-model with i defrees of freedom for the approach
    formula <- as.formula(paste("Surv(gsurv, gstatus) ~ ns(", covariate, ", df=", toString(j), ")"))
    aic[j] = AIC(coxph(formula, data=renaltx_clean))
  }
  
  # plot the aics for the splines, the lowest AIC is the best fit
  plot(aic, type="b", main=covariate, xlab="natural spline penalty (df)")
  # get the best pspline value 
  formula <- as.formula(paste("Surv(gsurv, gstatus) ~ pspline(", covariate, ", data=d)"))
  qq=summary(coxph(formula, data=renaltx_clean))
  psplinespvals[i] = qq$coefficients[2,6] # ? 
  # add a text with the positions at x and y (you can decipher that on your own (: )
  text(x=1, y=min(aic,na.rm=T)+(max(aic,na.rm=T)-min(aic,na.rm=T))/2, 
  paste("pspline nonlin p = ", round(psplinespvals[i],4)), adj=0, col=4, cex=0.9)
  
  # get the minimum AIC
  best_df = which.min(aic)
  # Test linearity by comparing AIcs
  # Linear model
  formula <- as.formula(paste("Surv(gsurv, gstatus) ~ ", covariate))
  linear_model <- coxph(formula, data = renaltx_clean)
  aic_linear <- AIC(linear_model)
  
  formula <- as.formula(paste("Surv(gsurv, gstatus) ~ ns(", covariate, ", df= ", toString(best_df), ")"))
  spline_model <- coxph(formula, data = renaltx_clean)
  aic_spline <- AIC(spline_model)
  
  # Compare AICs
  print(paste("AIC (Linear):", round(aic_linear, 2)))
  print(paste("AIC (Spline):", round(aic_spline, 2)))
  
  # Likelihood Ratio Test
  lrt <- anova(linear_model, spline_model, test = "Chisq")
  print(lrt)
  
  # Visual inspection
  #plot(spline_model)
}
covs_continu <- c("aantalre" , "acclft" ,"creat", "predias", "prac", "cregsh")

for(x in quant_vars) {
  plot_splines(x)
}
```
Based on the martingale residuals and splines we don't need transformations.

Although the acclft, and prac have better splines at >1 knot, looking at the difference of the value between the linear line and the smoothed lines AIC in combination with the Martingale, there is no reason to do anything extra.

```{r}
cox_model <- cph(surv_object ~ acclft + dgf + aantalre + creat + 
                  predias + uprotein + prac + cregsh,
                  data = renaltx_clean, x = TRUE, y = TRUE, surv = TRUE)
```
Next, checking for multicolinearity
```{r}
# VIF
vif_values <- vif(cox_model)
vif_values
```
Very acceptable VIF values. No concern for multicolinearity.

Let's move on to the next assuption of proportinoality in cox.
```{r}
ph_test <- cox.zph(cox_model)
# Schoenfeld residuals
ggcoxzph(ph_test)
ph_test
```
Based on the results, I would suggest applying Landmarking to acclft, dgf, and aantalre.

Create a new model based on these.

```{r}
# Compute concordance statistic (c-index)
# Using validate function with bootstrapping
set.seed(42)
validate_cox <- validate(cox_model, method = "boot", B = 200, dxy = TRUE)
print(validate_cox)

# c-index
dxy_value <- validate_cox["Dxy", "index.corrected"]
c_index <- (dxy_value / 2) + 0.5
print(c_index)


# Calibration plot
cal <- calibrate(cox_model, method = "boot", B = 200, u = 150)
plot(cal)

# time to event is minimum of 'gsurv' and 'psurv'?
renaltx_clean$time_to_event <- pmin(renaltx_clean$gsurv, renaltx_clean$psurv)

# 1 = graft rejection, 2 = death before graft rejection, 0 = censoring
renaltx_clean$event_type <- with(renaltx_clean,
                                 ifelse(gstatus == 1 & gsurv <= psurv, 1,
                                        ifelse(pstatus == 1 & psurv < gsurv, 2, 0)))

# Prepare covariate matrix exclude intercept from model.matrix
covariate_matrix <- model.matrix(~ acclft + dgf + aantalre + creat + predias +
                                   prac + uprotein + cregsh, data = renaltx_clean)[, -1]

# Fit Fine and Gray model for event type 1 (graft rejection)
fg_model <- crr(ftime = renaltx_clean$time_to_event,
                fstatus = renaltx_clean$event_type,
                cov1 = covariate_matrix,
                failcode = 1,
                cencode = 0)

summary(fg_model)
```
Compute best Landmarking breakpoints

```{r}
variables_landmarking <- c("acclft", "dgf", "aantalre")
results <- c()
for (var in variables_landmarking) {
  formula <- as.formula(paste("surv_object ~", var))
  # check out samples form 3-10 landmarks
  aics <- c()
  for (i in 1:10) {
    # get 5 intervals from 
    landmarks <- seq(min(renaltx_clean$gsurv), max(renaltx_clean$gsurv), length.out=i)
    # split data based on landmarks
    split = survSplit(formula, renaltx_clean, cut=landmarks, episode ="timegrp", start="tstart", end="tstop")
    
    landmark_formula <- as.formula(paste("surv_object ~", var, "* strata(timegrp)")) 
    cox_lm <- coxph(landmark_formula, data = split)
    aics[i] <- AIC(cox_lm)
  }
  plot(aics, main=paste("AICs for ", var))
}


# i will for no particular reason settle for 5

```

```{r}
# Split between landmarking and no landmarking
# all_variables <- c("acclft", "dgf", "aantalre", "creat", "predias", "prac", "uprotein", "cregsh")
variables_no_landmarking <- c("creat", "predias", "prac", "uprotein", "cregsh")
variables_landmarking <- c("acclft", "dgf", "aantalre")

univariable_results <- list()
# Univariate without landmarking
for (var in variables_no_landmarking) {
  formula <- as.formula(paste("surv_object ~", var))
  uni_cox <- coxph(formula, data = renaltx_clean)
  univariable_results[[var]] <- summary(uni_cox)
}

# Univariate with landmarking
lms <- seq(min(renaltx_clean$gsurv), max(renaltx_clean$gsurv), length.out=5)
# split data based on landmarks
for (var in variables_landmarking) {
  lm_formula <- as.formula(paste("surv_object ~", var))
  split_data = survSplit(lm_formula, renaltx_clean, cut=lms, episode ="timegrp", start="tstart", end="tstop")
  form <- as.formula(paste("surv_object ~", var, "* strata(timegrp)"))
  uni_cox <- coxph(form, data = split_data)
  univariable_results[[var]] <- summary(uni_cox)
}

for (var in names(univariable_results)) {
  cat("\nUnivariable Cox regression for", var, ":\n")
  print(univariable_results[[var]])
}
```

```{r}
# hazard ratios and confidence intervals

lms <- seq(min(renaltx_clean$gsurv), max(renaltx_clean$gsurv), length.out=5)
# split data based on landmarks
multi_formula <- as.formula(paste("surv_object ~ acclft + dgf + aantalre + creat +
                         predias + uprotein + prac + cregsh"))
multi_split_data = survSplit(multi_formula, renaltx_clean, cut=lms, episode ="timegrp", start="tstart", end="tstop")

library(rms)
dd <- datadist(multi_split_data)
options(datadist = "dd")

final_cox_model <- cph(surv_object ~ acclft + dgf + aantalre + rcs(creat, 4) +
                         rcs(predias, 4) + uprotein + prac + cregsh,
                       data = multi_split_data, x = TRUE, y = TRUE, surv = TRUE)


print(summary(final_cox_model))

# plot survival curves for different levels of 'dgf'
surv_fit <- survfit(final_cox_model, newdata = data.frame(
  acclft = median(renaltx_clean$acclft),
  dgf = levels(renaltx_clean$dgf),
  aantalre = median(as.numeric(as.character(renaltx_clean$aantalre))),
  creat = median(renaltx_clean$creat),
  predias = median(renaltx_clean$predias),
  prac = median(renaltx_clean$prac),
  uprotein = levels(renaltx_clean$uprotein)[1],
  cregsh = median(renaltx_clean$cregsh)
))

plot(surv_fit, col = 1:length(levels(renaltx_clean$dgf)),
     xlab = "Time (months)", ylab = "Survival Probability")
legend("bottomleft", legend = levels(renaltx_clean$dgf),
       col = 1:length(levels(renaltx_clean$dgf)), lty = 1)
```


