---
title: 'week 4 assignment: predictive value of repeated measures for graft and patient failure'
author: "Robin Manz, Matias Tolppanen"
date: "2024-11-21"
output: html_notebook
---

## Setup 

```{r setup}
# load librarys
library(ggplot2)
library(survival)
library(nlme)
library(JM)
```
## Loaddata

```{r}
load(file="kidney_transplant.RData")
ls()

```
## Descriptives
the object "d" contains data of 838 patients who received a kidney transplant for end-stage renal failure and contains 40 columns, including

| column | description | 
| --- | --- | 
| ID | patient ID number (1,2,3,...,838) |
| time_to_graft_failure | years from transplantation till date of graft failure |
| time_to_death | years from transplantation till date of death |
| stat_pat  | 1/0 (1 = patient died; 0 = patient is alive at last followup) |
| stat_gra |  1/0 (1 = graft failed; 0 = graft still functioning at last fup) | 
| --- |--- | 
| sex_pat | 1 female, 2= male | 
| age_at_tx | age at transplatation in years | 
| type_dia | dialysis type: CAPD, geen_dialyse, Hemodialyse centrum, Hemodialyse thuis |
| duur_dia | Dialysis duration Time, Integer |
| bmi | float BMI | 
| sexdon | Sex of Donor (1 = female, 2 = male) | 
| agedon | Age of Donor, Integer | 
| retrans | Transplatiation type (0 = first transplatation, 1 = retransplantation) | 
| gfr(1-7) | time dependeant, gluomerular filtration rate | 
| screat(1-7) | time dependeant, createnine levels | 
| map(1-7) | time dependeant, mid arterial pressure | 
| (Time periods) | 3m, 6m, 1y, 2y, 5y, 10y, 15y | 

 The object "dlong" contains the same data but each patient is now represented by seven records, one for each repeated measure.
 
 
 **Question**: The general aim of the assignment is to develop dynamic prediction models for graft failure and death using both the baseline characteristics and the repeatedly measured covariates.
Use both Cox regression models including the time-dependent covariates gfr, screat and map and joint Cox-lme models with the same data. Discuss the different assumptions involved.

## Descriptives 
```{r}
# Basic numbers
n_entries <- nrow(d)
n_f <- nrow(subset(d, sex_pat == "female"))
n_m <- nrow(subset(d, sex_pat == "male"))

print(paste("Total patients:", n_entries, "female:", n_f, ", male:", n_m))
sapply(d, function(x) sum(is.na(x)))


print(paste("Number of entries", n_entries))
n_graft_rejection <- nrow(subset(d, stat_gra == "Graft loss"))
print(paste("Number of Graft_rejections", n_graft_rejection))
n_died <- nrow(subset(d, stat_pat == "overleden"))
print(paste("Number of Deaths", n_died))


# histograms of continuous variables
hist(d$age_at_tx, main="Age at transplanation")
hist(d$duur_dia, main="Duration of dialysis")
hist(d$bmi, breaks=20, main="BMI")
hist(d$agedon, main="Age of Donor")

# creat, gfr, map
hist(dlong$creat, main="Histogram for Creat")
creat_mean <- mean(dlong$creat ,na.rm = TRUE)
creat_sd <- sd(dlong$creat, na.rm = TRUE)
creat_min <- min(dlong$creat, na.rm = TRUE)
creat_max <- max(dlong$creat, na.rm = TRUE)
print(paste("Createnine mean:", round(creat_mean, 2), "(SD =", round(creat_sd, 2), "), min =", round(creat_min, 2), ", max = ", round(creat_max, 2)))
# gfr
hist(dlong$gfr, main="Histogram for gfr")
gfr_mean <- mean(dlong$gfr, na.rm = TRUE)
gfr_sd<-sd(dlong$gfr, na.rm = TRUE)
gfr_min<-min(dlong$gfr, na.rm = TRUE)
gfr_max<-max(dlong$gfr, na.rm = TRUE)
print(paste("Gfr mean:", round(gfr_mean, 2), "(SD =", round(gfr_sd, 2), "), min =", round(gfr_min, 2), ", max = ", round(gfr_max, 2)))
# map
hist(dlong$map, main="Histogram for map")
map_mean<-mean(dlong$map, na.rm = TRUE)
map_sd<-sd(dlong$map, na.rm = TRUE)
map_min<-min(dlong$map, na.rm = TRUE)
map_max<-max(dlong$map, na.rm = TRUE)
print(paste("map mean:", round(map_mean, 2), "(SD =", round(map_sd, 2), "), min =", round(map_min, 2), ", max = ", round(map_max)))


```
## add start and stop to the dlong for processing
```{r}
# really bad appriach i know ....
# create start and stop events in the dataset
start <- numeric(nrow(dlong))
stop <- numeric(nrow(dlong))
counter = 1
for (row in 1:nrow(dlong)) {
  if (counter == 8) {
    counter <- 1
  }
  if (counter == 1) {
    # new patient start = 0
    start[row] <- 0 
    print(dlong[row, "years"])
    stop[row] <- dlong[row, "years"] 
  } else {
    # existing patient start = last stop
    start[row] <- stop[row - 1]
    print(dlong[row, "years"])
    stop[row] <- dlong[row, "years"]
  }
  counter <- counter + 1
}
dlong$start <- start
dlong$stop <- stop


```
## Kaplan meier for graft rejection 

````{r}

Y <- Surv(time = d$time_to_graft_failure, event = d$stat_gra == "Graft loss")
summary(Y)

# plot the Surv time
plot(Y, lty = c("solid", "dashed", "dashed"), col = c("black", "grey", "grey"), xlab = "Survival Time In Years", ylab = "Graft rejection probabilities for all patients", main="Kaplan-Meier Curve")


```
## Preprocessing , checking assumptions etc. 

```{r}
# get subset of only graft rejected patients
d_graft_loss <- subset(d, stat_gra == "Graft loss")

# check for linearity and proportionality
# check for linearity with psplines splines for each continuous variable
cox_m <- coxph(Surv(time = d$time_to_graft_failure, event = d$stat_gra == "Graft loss") ~ sex_pat + pspline(age_at_tx) + type_dia + pspline(duur_dia) + pspline(bmi) + sexdon + pspline(agedon) + retrans,  data = d, x = TRUE)
summary(cox_m)
# all variables are linear

# check for proportionality multivariate model
cox_p <- coxph(Surv(time = d$time_to_graft_failure, event = d$stat_gra == "Graft loss") ~ sex_pat + age_at_tx + type_dia + duur_dia + bmi + sexdon + agedon + retrans,  data = d, x = TRUE)
summary(cox_p)
cox_zph=cox.zph(cox_p)
plot(cox_zph[1])
plot(cox_zph[2])
plot(cox_zph[3])
plot(cox_zph[4])
plot(cox_zph[5])
plot(cox_zph[6])
plot(cox_zph[7])
plot(cox_zph[8])
print(cox_zph)

# TODO check univariate?

# Univariate check for predictive power
print("##################Death ####################")
# variable selection for death
summary(coxph(Surv(time_to_death,as.numeric(stat_pat)-1)~sex_pat,data=d))
summary(coxph(Surv(time_to_death,as.numeric(stat_pat)-1)~age_at_tx,data=d))
summary(coxph(Surv(time_to_death,as.numeric(stat_pat)-1)~bmi,data=d))
summary(coxph(Surv(time_to_death,as.numeric(stat_pat)-1)~sexdon,data=d))
summary(coxph(Surv(time_to_death,as.numeric(stat_pat)-1)~agedon,data=d))
summary(coxph(Surv(time_to_death,as.numeric(stat_pat)-1)~type_dia,data=d))
summary(coxph(Surv(time_to_death,as.numeric(stat_pat)-1)~duur_dia,data=d))
summary(coxph(Surv(time_to_death,as.numeric(stat_pat)-1)~retrans,data=d))

print("############### Graft failure ################")
# variable selection for graft failire
summary(coxph(Surv(time_to_graft_failure,as.numeric(stat_gra)-1)~sex_pat,data=d))
summary(coxph(Surv(time_to_graft_failure,as.numeric(stat_gra)-1)~age_at_tx,data=d))
summary(coxph(Surv(time_to_graft_failure,as.numeric(stat_gra)-1)~bmi,data=d))
summary(coxph(Surv(time_to_graft_failure,as.numeric(stat_gra)-1)~sexdon,data=d))
summary(coxph(Surv(time_to_graft_failure,as.numeric(stat_gra)-1)~agedon,data=d))
summary(coxph(Surv(time_to_graft_failure,as.numeric(stat_gra)-1)~type_dia,data=d))
summary(coxph(Surv(time_to_graft_failure,as.numeric(stat_gra)-1)~duur_dia,data=d))
summary(coxph(Surv(time_to_graft_failure,as.numeric(stat_gra)-1)~retrans,data=d))


```
## cox cph Graft rejection variables
sig: age_at_tx (2.6e-10), agedon (6.53e-05),  type_dia (cebtrzm, 0.0198), 
non-sig: sex_pat(close 0.0556), bmi, sexdon, duur_dia, retrans

## coxacph Death variables
significant predictors: 
- duur_dia (0.000423), agedon (0.00998),  age_at_tx (<2e-16), type_diaHemodialyse centrum (0.00929 )
non-significant: sex_pat, retrans, sexdon, bmi


## create linear mixed effect model 
```{r}
lme_creat_model <- lme(creat ~ gfr + creat + map + sex_pat + age_at_tx + bmi + type_dia+ duur_dia + sexdon + agedon + retrans, data = dlong)


# baseline_model <- lmer(diameter ~ 1 + (1|patnr), data = data )
```
## Check for linarity using pslines:
significant for linear: 
- age_at_tx
- duur_dia
- bmi
- agedon

## Proportionality 
- sex_pat ok, -1 < x <1
- age_at_tx ok, -0.1 < x < 0.1
- type_dia p-values says ok, (proportionality violated)
- duur_dia ok
- bmi ok
- sex_don not ok
- age_don ok
- retrans ok


## Cox models
```{r}
# cox model greaft failure
m1 <- coxph(Surv(time_to_graft_failure, as.numeric(stat_gra)-1) ~ age_at_tx + type_dia + agedon,data = d, x = TRUE)

# for death
m1 <- coxph(Surv(time_to_death,as.numeric(stat_pat)-1)~ age_at_tx + agedon + type_dia +
         duur_dia , data=d, x=TRUE)

m1
# lme model
s1 <- 
```
  
## Joint model Cox-lme for graft failure 
```{r}
# joint model
j1 <- jointModel(s2, m1, timeVar = "obstime")

```