---
title: 'Week 3 assignment: Predictive value of repeated measures for graft and patient failure'
author: "Robin Manz, Matias Tolppanen"
date: "2024-11-22"
output: html_notebook
---

```{r}
library(tidyverse)
library(nlme)
library(survival)
library(JM)
library(ggpubr)
library(mice)
library(VIM)
library(e1071)
library(dplyr)
library(tidyr)
library(gridExtra)

```

```{r}
load('kidney_transplant.RData')
ls()
```

Data description and exploration

```{r}
str(d)
str(dlong)

summary(d)
summary(dlong)

sapply(d, class)
sapply(dlong, class)
```

There seems to be outliers on gfr and creat that are clinically impossible. Let's investigate.

```{r}
limits <- list(
  gfr = c(0, 150),
  creat = c(0, 750),
  map = c(0, 170)
)


# Create a function to detect and remove outliers
handle_outliers <- function(data, variable, limits) {
  lower_limit <- limits[1]
  upper_limit <- limits[2]
  
  outliers <- which(data[[variable]] < lower_limit | data[[variable]] > upper_limit)
  n_outliers <- length(outliers)
  
  cat(sprintf("Variable '%s': %d outliers detected (outside [%g, %g])\n", 
              variable, n_outliers, lower_limit, upper_limit))
  
  if (n_outliers > 0) {
    data <- data[-outliers, ]
  }
  
  return(data)
}

continuous_vars <- c("gfr", "creat", "map")
for (var in continuous_vars) {
  dlong <- handle_outliers(dlong, var, limits[[var]])
}

cat("\nSummary of cleaned variables:\n")
summary(dlong[, continuous_vars])
```

Missing data

```{r}
aggr_plot <- aggr(d, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE,
                  labels=names(d), cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))

aggr_plot_long <- aggr(dlong, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE,
                       labels=names(dlong), cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))

# Proportion of missing data
colSums(is.na(d)) / nrow(d)
colSums(is.na(dlong)) / nrow(dlong)

```

missingness seems random and minimal/non-significant. For many of the measurements we can expect high missigness due to the nature of that data. We shall proceed with complete-case analysis

```{r}
# Variables to explore
baseline_vars <- c('sex_pat', 'age_at_tx', 'type_dia', 'duur_dia', 'bmi', 'sexdon', 'agedon', 'retrans')


d %>% ggplot(aes(x = age_at_tx)) + 
  geom_histogram(binwidth = 5, fill = 'blue', color = 'black') + 
  theme_minimal() + 
  labs(title = 'Distribution of Age at Transplantation', x = 'Age', y = 'Count')

d %>% ggplot(aes(x = bmi)) + 
  geom_histogram(binwidth = 1, fill = 'green', color = 'black') + 
  theme_minimal() + 
  labs(title = 'Distribution of BMI', x = 'BMI', y = 'Count')

d %>% ggplot(aes(x = duur_dia)) + 
  geom_histogram(binwidth = 5, fill = 'red', color = 'black') + 
  theme_minimal() + 
  labs(title = 'Distribution of Duration of Dialysis', x = 'Duration (months)', y = 'Count')

d %>% ggplot(aes(x = sex_pat)) + 
  geom_bar(fill = 'orange') + 
  theme_minimal() + 
  labs(title = 'Distribution of Patient Sex', x = 'Sex', y = 'Count')

d %>% ggplot(aes(x = type_dia)) + 
  geom_bar(fill = 'purple') + 
  theme_minimal() + 
  labs(title = 'Distribution of Dialysis Type', x = 'Type of Dialysis', y = 'Count')

d %>% ggplot(aes(x = sexdon)) + 
  geom_bar(fill = 'cyan') + 
  theme_minimal() + 
  labs(title = 'Distribution of Donor Sex', x = 'Sex', y = 'Count')

d %>% ggplot(aes(x = factor(retrans))) + 
  geom_bar(fill = 'pink') + 
  theme_minimal() + 
  labs(title = 'Distribution of Re-transplantation', x = 'Re-transplantation (1=Yes, 0=No)', y = 'Count')

## Outcome Variables
d %>% ggplot(aes(x = stat_gra)) + 
  geom_bar(fill = 'gray') + 
  theme_minimal() + 
  labs(title = 'Graft Failure Status', x = 'Graft Status', y = 'Count')

d %>% ggplot(aes(x = stat_pat)) + 
  geom_bar(fill = 'brown') + 
  theme_minimal() + 
  labs(title = 'Patient Survival Status', x = 'Patient Status', y = 'Count')

d %>% ggplot(aes(x = time_to_graft_failure)) + 
  geom_histogram(binwidth = 1, fill = 'steelblue', color = 'black') + 
  theme_minimal() + 
  labs(title = 'Time to Graft Failure', x = 'Years', y = 'Count')

d %>% ggplot(aes(x = time_to_death)) + 
  geom_histogram(binwidth = 1, fill = 'darkgreen', color = 'black') + 
  theme_minimal() + 
  labs(title = 'Time to Death', x = 'Years', y = 'Count')


dlong %>% ggplot(aes(x = gfr)) + 
  geom_histogram(binwidth = 5, fill = 'lightblue', color = 'black') + 
  theme_minimal() + 
  labs(title = 'Distribution of GFR', x = 'GFR', y = 'Count')

dlong %>% ggplot(aes(x = creat)) + 
  geom_histogram(binwidth = 10, fill = 'lightgreen', color = 'black') + 
  theme_minimal() + 
  labs(title = 'Distribution of Creatinine', x = 'Creatinine', y = 'Count')

dlong %>% ggplot(aes(x = map)) + 
  geom_histogram(binwidth = 5, fill = 'lightpink', color = 'black') + 
  theme_minimal() + 
  labs(title = 'Distribution of MAP', x = 'Mean Arterial Pressure', y = 'Count')

dlong %>% ggplot(aes(x = years, y = gfr)) + 
  geom_point(alpha = 0.3) + 
  geom_smooth(method = 'loess') +
  theme_minimal() + 
  labs(title = 'GFR over Time', x = 'Years since Transplantation', y = 'GFR')

dlong %>% ggplot(aes(x = years, y = creat)) + 
  geom_point(alpha = 0.3) + 
  geom_smooth(method = 'loess') +
  theme_minimal() + 
  labs(title = 'Creatinine over Time', x = 'Years since Transplantation', y = 'Creatinine')

dlong %>% ggplot(aes(x = years, y = map)) + 
  geom_point(alpha = 0.3) + 
  geom_smooth(method = 'loess') +
  theme_minimal() + 
  labs(title = 'MAP over Time', x = 'Years since Transplantation', y = 'Mean Arterial Pressure')
```

### Distribution Checks for Continuous Variables

```{r}
# List of continuous variables
continuous_vars <- c('gfr', 'creat', 'map')

# Histograms for 'dlong' continuous variables
# List of continuous variables
continuous_vars <- c('gfr', 'creat', 'map')

# Histograms for 'dlong' continuous variables
for (var in continuous_vars) {
  print(
    ggplot(dlong, aes_string(x = var)) +
      geom_histogram(binwidth = 5, fill = 'blue', color = 'black', na.rm = TRUE) +
      theme_minimal() +
      labs(title = paste('Distribution of', var), x = var, y = 'Count') +
      geom_vline(xintercept = mean(dlong[[var]], na.rm = TRUE), color = 'red', linetype = 'dashed') +
      theme(plot.title = element_text(hjust = 0.5))
  )
}


# Boxplots for 'dlong' continuous variables
for (var in continuous_vars) {
  print(
    ggplot(dlong, aes_string(y = var)) +
      geom_boxplot(fill = 'orange', color = 'black', na.rm = TRUE) +
      theme_minimal() +
      labs(title = paste('Boxplot of', var), y = var) +
      theme(plot.title = element_text(hjust = 0.5))
  )
  print(
    ggplot(dlong, aes_string(sample=var)) +
      stat_qq() +
      stat_qq_line() +
      labs(title = paste('Q-Q Plot of', var), y = var)
  )
}


```

Based on the graphs, I think creat would definetely benefit from log transform.

For completeness, let's check skeweness numerically

```{r, warning=FALSE}
for (var in continuous_vars) {
  skewness_value <- skewness(dlong[[var]], na.rm = TRUE)
  cat(paste('Skewness of', var, ':', round(skewness_value, 2)), '\n')
}
```

Creat definitely needs log transform.

```{r, warning=FALSE}
ggplot(dlong, aes(x = creat)) +
  geom_histogram(binwidth = 10, fill = 'blue', color = 'black') +
  theme_minimal() +
  labs(title = 'Creatinine Before Transformation', x = 'Creatinine', y = 'Count')

# Apply log transformation
dlong <- dlong %>% mutate(creat = log(creat))

# Histogram after transformation
ggplot(dlong, aes(x = creat)) +
  geom_histogram(binwidth = 0.1, fill = 'green', color = 'black') +
  theme_minimal() +
  labs(title = 'Creatinine After Log Transformation', x = 'Log(Creatinine)', y = 'Count')
```

Way better.

### Next well choose the best **linear xixed-effects models**

```{r, warning=FALSE}
lme_gfr_intercept <- lme(gfr ~ years, data = dlong, random = ~ 1 | ID, na.action = na.exclude)
summary(lme_gfr_intercept)

```

```{r}
lme_gfr_slope <- lme(gfr ~ years, data = dlong, random = ~ years | ID, na.action = na.exclude)
summary(lme_gfr_slope)

# Extract fixed effects
fixed_effects <- summary(lme_gfr_slope)$tTable

# Create a data frame for the table
fixed_table <- data.frame(
    Variable = rownames(fixed_effects),
    # Estimate = fixed_effects[, "Value"],
    `Std. Error` = fixed_effects[, "Std.Error"],
    `t-value` = fixed_effects[, "t-value"],
    `p-value` = fixed_effects[, "p-value"]
)

# View the table
print(fixed_table)

# Extract random effects
random_effects <- VarCorr(lme_gfr_slope)

# Create a data frame for the table
random_table <- data.frame(
    `Random Effect` = c("Intercept", "Years", "Residual"),
    `Variance` = c(as.numeric(random_effects[1, 1]), 
                   as.numeric(random_effects[2, 2]), 
                    summary(lme_gfr_slope)$sigma^2),
    `Standard Deviation` = c(sqrt(as.numeric(random_effects[1, 1])), 
                             sqrt(as.numeric(random_effects[2, 2])), 
                             summary(lme_gfr_slope)$sigma)
)

# View the table
print(random_table)

```

Comparing using aic and bic

```{r}
gfr_model_comparison <- data.frame(
  Model = c("Random Intercept", "Random Intercept + Slope"),
  AIC = c(AIC(lme_gfr_intercept), AIC(lme_gfr_slope)),
  BIC = c(BIC(lme_gfr_intercept), BIC(lme_gfr_slope))
)
gfr_model_comparison

anova(lme_gfr_intercept, lme_gfr_slope)

```

Random Intercept + Slope + Quadratic seems best.

### Creat lme

```{r, warning=FALSE}
lme_creat_intercept <- lme(creat ~ years, data = dlong, random = ~ 1 | ID, na.action = na.exclude)
summary(lme_creat_intercept)


```

```{r}
lme_creat_slope <- lme(creat ~ years, data = dlong, random = ~ years | ID, na.action = na.exclude)
summary(lme_creat_slope)

# Extract fixed effects
fixed_effects <- summary(lme_creat_slope)$tTable

# Create a data frame for the table
fixed_table <- data.frame(
    Variable = rownames(fixed_effects),
    Estimate = fixed_effects[, "Value"],
    `Std. Error` = fixed_effects[, "Std.Error"],
    `t-value` = fixed_effects[, "t-value"],
    `p-value` = fixed_effects[, "p-value"]
)

# View the table
print(fixed_table)

# Extract random effects
random_effects <- VarCorr(lme_creat_slope)

# Create a data frame for the table
random_table <- data.frame(
    `Random Effect` = c("Intercept", "Years", "Residual"),
    `Variance` = c(as.numeric(random_effects[1, 1]), 
                   as.numeric(random_effects[2, 2]), 
                    summary(lme_creat_slope)$sigma^2),
    `Standard Deviation` = c(sqrt(as.numeric(random_effects[1, 1])), 
                             sqrt(as.numeric(random_effects[2, 2])), 
                             summary(lme_creat_slope)$sigma)
)

# View the table
print(random_table)

```

```{r}
creat_model_comparison <- data.frame(
  Model = c("Random Intercept", "Random Intercept + Slope"),
  AIC = c(AIC(lme_creat_intercept), AIC(lme_creat_slope)),
  BIC = c(BIC(lme_creat_intercept), BIC(lme_creat_slope))
)
creat_model_comparison

anova(lme_creat_intercept, lme_creat_slope)

```

+:------------------------------------------------+
| Random Intercept + Slope + Quadratic seems best |
|                                                 |
| ### Map                                         |
+-------------------------------------------------+

```{r}
lme_map_intercept <- lme(map ~ years, data = dlong, random = ~ 1 | ID, na.action = na.exclude)
summary(lme_map_intercept)


```

```{r}
lme_map_slope <- lme(map ~ years, data = dlong, random = ~ years | ID, na.action = na.exclude)
summary(lme_map_slope)

# Extract fixed effects
fixed_effects <- summary(lme_map_slope)$tTable

# Create a data frame for the table
fixed_table <- data.frame(
    Variable = rownames(fixed_effects),
    Estimate = fixed_effects[, "Value"],
    `Std. Error` = fixed_effects[, "Std.Error"],
    `t-value` = fixed_effects[, "t-value"],
    `p-value` = fixed_effects[, "p-value"]
)

# View the table
print(fixed_table)

# Extract random effects
random_effects <- VarCorr(lme_map_slope)

# Create a data frame for the table
random_table <- data.frame(
    `Random Effect` = c("Intercept", "Years", "Residual"),
    `Variance` = c(as.numeric(random_effects[1, 1]), 
                   as.numeric(random_effects[2, 2]), 
                   # summary(lme_map_slope)$sigma^2),
    `Standard Deviation` = c(sqrt(as.numeric(random_effects[1, 1])), 
                             sqrt(as.numeric(random_effects[2, 2])), 
                             summary(lme_map_slope)$sigma)
)

# View the table
print(random_table)

```

```{r}
map_model_comparison <- data.frame(
  Model = c("Random Intercept", "Random Intercept + Slope"),
  AIC = c(AIC(lme_map_intercept), AIC(lme_map_slope)),
  BIC = c(BIC(lme_map_intercept), BIC(lme_map_slope))
)
map_model_comparison

anova(lme_map_intercept, lme_map_slope)

```

Random Intercept + Slope seems best

### Next, cox. We'll choose the best parameters using step wise AIC
## Graft rejection
```{r}
# Create event indicator in 'd' (for survival model)
d$event <- ifelse(d$stat_gra == "Graft loss", 1, 0)

# potential predictors (exclude 'ID' from predictors)
cox_vars <- c("age_at_tx", "sex_pat", "bmi", "type_dia", "duur_dia", "sexdon", "agedon", "retrans")

cox_data <- d[, c("ID", "time_to_graft_failure", "event", cox_vars)]

cox_full <- coxph(Surv(time_to_graft_failure, event) ~ . - ID, data = cox_data, x = TRUE)
summary(cox_full)

```

Analog for death
```{r}
# Create event indicator in 'd' (for survival model)
d$event_death <- ifelse(d$stat_pat == "overleden", 1, 0)

# potential predictors (exclude 'ID' from predictors)
cox_death_vars <- c("age_at_tx", "sex_pat", "bmi", "type_dia", "duur_dia", "sexdon", "agedon", "retrans")

cox_death_data <- d[, c("ID", "time_to_death", "event_death", cox_vars)]

cox_death_full <- coxph(Surv(time_to_death, event_death) ~ . - ID, data = cox_death_data, x = TRUE)
summary(cox_death_full)


```

# Performing backward stepwise selection

```{r}
cox_selected <- stepAIC(cox_full, direction = "backward")
summary(cox_selected)

cox_death_selected <- stepAIC(cox_death_full, direction = "backward")
summary(cox_death_selected)
```

## Cox with the selected parameters

```{r}
cox_selected <- update(cox_selected, . ~ . + cluster(ID), x = TRUE, data = cox_data)
summary(cox_selected)

# death as well
cox_death_selected <- update(cox_death_selected, . ~ . + cluster(ID), x = TRUE, data = cox_death_data)
summary(cox_death_selected)

```
Significant for graft rejection: age_at_tx,sex_patmale, type_diageen, agedon

Significant for death: age_at_tx, duur_dia
mildly: agedon (0.097), retrans (0.069297)

### Joint Model for GFR

```{r}
joint_model_gfr <- jointModel(lme_gfr_slope, cox_selected, timeVar = "years")
summary(joint_model_gfr)

death_joint_model_gfr <- jointModel(lme_gfr_slope, cox_death_selected, timeVar = "years")
summary(death_joint_model_gfr)
```

### Joint Model for Creatine

```{r}
joint_model_creat <- jointModel(lme_creat_slope, cox_selected, timeVar = "years")
summary(joint_model_creat)

death_joint_model_creat <- jointModel(lme_creat_slope, cox_death_selected, timeVar = "years")
summary(death_joint_model_creat)

```

### Joint Model for MAP

```{r}
joint_model_map <- jointModel(lme_map_slope, cox_selected, timeVar = "years")
summary(joint_model_map)

death_joint_model_map <- jointModel(lme_map_slope, cox_death_selected, timeVar = "years")
summary(death_joint_model_map)
```

## Model Comparison

```{r}
joint_model_comparison <- data.frame(
  Model = c("Joint Model GFR", "Joint Model Creatinine", "Joint Model MAP"),
  AIC = c(AIC(joint_model_gfr), AIC(joint_model_creat), AIC(joint_model_map))
)
joint_model_comparison

death_joint_model_comparison <- data.frame(
  Model = c("Death Joint Model GFR", "Death Joint Model Creatinine", "Death Joint Model MAP"),
  AIC = c(AIC(death_joint_model_gfr), AIC(death_joint_model_creat), AIC(death_joint_model_map))
)
death_joint_model_comparison

```

## Creatine seems the absolute best

```{r}
plot(joint_model_gfr, which = 2, type = "p")
plot(joint_model_creat, which = 2, type = "p")
plot(joint_model_map, which = 2, type = "p")

plot(death_joint_model_gfr, which = 2, type = "p")
plot(death_joint_model_creat, which = 2, type = "p")
plot(death_joint_model_map, which = 2, type = "p")
```

## Splitting the data into training and test sets

```{r}
set.seed(42)
train_ids <- sample(unique(d$ID), size = 0.8 * length(unique(d$ID)))
test_ids <- setdiff(unique(d$ID), train_ids)

d_train <- d %>% filter(ID %in% train_ids)
dlong_train <- dlong %>% filter(ID %in% train_ids)

d_test <- d %>% filter(ID %in% test_ids)
dlong_test <- dlong %>% filter(ID %in% test_ids)

```

# lme with training data

```{r}
lme_gfr_slope_train <- lme(gfr ~ years, data = dlong_train, random = ~ years | ID, na.action = na.exclude)

lme_creat_slope_train <- lme(creat ~ years, data = dlong_train, random = ~ years | ID, na.action = na.exclude)

lme_map_slope_train <- lme(map ~ years, data = dlong_train, random = ~ years | ID, na.action = na.exclude)

```

```{r}
# 'event' defined in training data
d_train$event <- ifelse(d_train$stat_gra == "Graft loss", 1, 0)

# Re-fit Cox model with the before selected predictors
cox_selected_train <- coxph(formula(cox_selected), data = d_train, x = TRUE)
summary(cox_selected_train)

# 'event' defined in training data
d_train$event_death <- ifelse(d_train$stat_pat == "overleden", 1, 0)

# Re-fit Cox model with the before selected predictors
cox_death_selected_train <- coxph(formula(cox_death_selected), data = d_train, x = TRUE)
summary(cox_death_selected_train)

```

## Joint Train models

```{r}
joint_model_gfr_train <- jointModel(lme_gfr_slope_train, cox_selected_train, timeVar = "years")

joint_model_creat_train <- jointModel(lme_creat_slope_train, cox_selected_train, timeVar = "years")

joint_model_map_train <- jointModel(lme_map_slope_train, cox_selected_train, timeVar = "years")


death_joint_model_gfr_train <- jointModel(lme_gfr_slope_train, cox_death_selected_train, timeVar = "years")

death_joint_model_creat_train <- jointModel(lme_creat_slope_train, cox_death_selected_train, timeVar = "years")

death_joint_model_map_train <- jointModel(lme_map_slope_train, cox_death_selected_train, timeVar = "years")

```

## Dynamic predictions and plots

3 test patients.

```{r}
# Choose 3 test patient IDs
set.seed(42)
test_patient_ids <- sample(test_ids, 3)
test_patient_ids

generate_patient_plots <- function(patient_id) {
  test_patient_data <- dlong_test %>% filter(ID == patient_id)

  test_patient_data$creat <- log(test_patient_data$creat)

  patient_data_gfr <- test_patient_data %>% filter(!is.na(gfr))
  patient_data_creat <- test_patient_data %>% filter(!is.na(creat))
  patient_data_map <- test_patient_data %>% filter(!is.na(map))

  dyn_pred_gfr <- survfitJM(joint_model_gfr_train, newdata = patient_data_gfr, type = "Subject", idVar = "ID", interval = "confidence", last.time = max(patient_data_gfr$years))
  
  plot(dyn_pred_gfr, main=paste("Graft rejection probability for", patient_id, "using gfr"), conf.int=TRUE, fill.area=TRUE, col.area="grey", include.y=TRUE)
  
  dyn_pred_creat <- survfitJM(joint_model_creat_train, newdata = patient_data_creat, type = "Subject", idVar = "ID", interval = "confidence", last.time = max(patient_data_creat$years))
  
  plot(dyn_pred_creat, main=paste("Graft rejection probability for", patient_id, "using creat"), conf.int=TRUE, fill.area=TRUE, col.area="grey", include.y=TRUE)
  
  dyn_pred_map <- survfitJM(joint_model_map_train, newdata = patient_data_map, type = "Subject", idVar = "ID", interval = "confidence", last.time = max(patient_data_map$years))
  
  plot(dyn_pred_map, main=paste("Graft rejection probability for", patient_id, "using map"), conf.int=TRUE, fill.area=TRUE, col.area="grey", include.y=TRUE)
}


death_generate_patient_plots <- function(patient_id) {
  test_patient_data <- dlong_test %>% filter(ID == patient_id)

  test_patient_data$creat <- log(test_patient_data$creat)

  patient_data_gfr <- test_patient_data %>% filter(!is.na(gfr))
  patient_data_creat <- test_patient_data %>% filter(!is.na(creat))
  patient_data_map <- test_patient_data %>% filter(!is.na(map))

  dyn_pred_gfr <- survfitJM(death_joint_model_gfr_train, newdata = patient_data_gfr, type = "Subject", idVar = "ID", interval = "confidence", last.time = max(patient_data_gfr$years))
  
  plot(dyn_pred_gfr, main=paste("Death probability for", patient_id, "using gfr"), conf.int=TRUE, fill.area=TRUE, col.area="grey", include.y=TRUE)
  
  dyn_pred_creat <- survfitJM(death_joint_model_creat_train, newdata = patient_data_creat, type = "Subject", idVar = "ID", interval = "confidence", last.time = max(patient_data_creat$years))
  
  plot(dyn_pred_creat, main=paste("Death probability for", patient_id, "using creat"), conf.int=TRUE, fill.area=TRUE, col.area="grey", include.y=TRUE)
  
  dyn_pred_map <- survfitJM(death_joint_model_map_train, newdata = patient_data_map, type = "Subject", idVar = "ID", interval = "confidence", last.time = max(patient_data_map$years))
  
  plot(dyn_pred_map, main=paste("Death probability for", patient_id, "using map"),  conf.int=TRUE, fill.area=TRUE, col.area="grey", include.y=TRUE)
}


for (patient_id in test_patient_ids) {
  print(paste("Graft rejection for", patient_id))
  generate_patient_plots(patient_id)
  print(paste("Death for", patient_id))
  death_generate_patient_plots(patient_id)
}

```
