---
title: "MAM03 Week 3 Aorta Diameter Growth in Marfan Syndrome"
output: html_document
date: '2024-11-13'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Authors: Robin Manz & Matias Tolppanen

```{r}
# Load necessary packages
library(ggplot2)
library(lme4)
library(dplyr)
library(boot)
```

```{r}
marfan <- read.csv("marfan.csv")
```

# 1. Data Description
# --------------------
```{r}
marfan$sexe <- factor(marfan$sexe, levels = c(0, 1), labels = c("Female", "Male"))
head(marfan)
cat("The dataset contains", nrow(marfan), "observations of", ncol(marfan), "variables.\n")
str(marfan)
```
Some summary statistics for numeric variables
```{r}
numeric_vars <- dplyr::select(marfan, age, diameter)
summary(numeric_vars)

descriptive_stats <- marfan %>%
  summarise(
    `Number of Patients` = n_distinct(patnr),
    `Total Measurements` = n(),
    `Age (Mean)` = mean(age),
    `Age (SD)` = sd(age),
    `Age (Min)` = min(age),
    `Age (Max)` = max(age),
    `Diameter (Mean)` = mean(diameter),
    `Diameter (SD)` = sd(diameter),
    `Diameter (Min)` = min(diameter),
    `Diameter (Max)` = max(diameter)
  )

descriptive_stats
```
```{r}
# statistics by gender
descriptive_by_gender <- marfan %>%
  group_by(sexe) %>%
  summarise(
    `Number of Patients` = n_distinct(patnr),
    `Total Measurements` = n(),
    `Age (Mean)` = mean(age),
    `Age (SD)` = sd(age),
    `Age (Min)` = min(age),
    `Age (Max)` = max(age),
    `Diameter (Mean)` = mean(diameter),
    `Diameter (SD)` = sd(diameter),
    `Diameter (Min)` = min(diameter),
    `Diameter (Max)` = max(diameter)
  )

descriptive_by_gender
```
```{r}
# n measurements per patient
measurements_per_patient <- marfan %>%
  group_by(patnr) %>%
  summarise(`Number of Measurements` = n())

# measurements per patient
measurements_summary <- measurements_per_patient %>%
  summarise(
    `Patients` = n(),
    `Measurements (Mean)` = mean(`Number of Measurements`),
    `Measurements (SD)` = sd(`Number of Measurements`),
    `Measurements (Min)` = min(`Number of Measurements`),
    `Measurements (Max)` = max(`Number of Measurements`)
  )

measurements_summary

ggplot(measurements_per_patient, aes(x = `Number of Measurements`)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
  labs(title = "Histogram of Number of Measurements per Patient",
       x = "Number of Measurements", y = "Count") +
  theme_minimal()

gender_distribution <- marfan %>%
  group_by(patnr) %>%
  summarise(sexe = first(sexe)) %>%
  count(sexe)

gender_distribution

ggplot(gender_distribution, aes(x = sexe, y = n, fill = sexe)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Patients by Gender",
       x = "Gender", y = "Number of Patients") +
  theme_minimal() +
  guides(fill = FALSE)

# Histograms of age and diameter by gender
ggplot(marfan, aes(x = age, fill = sexe)) +
  geom_histogram(binwidth = 2, position = "dodge", color = "black") +
  labs(title = "Histogram of Age by Gender", x = "Age (years)", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "lightblue"))

ggplot(marfan, aes(x = diameter, fill = sexe)) +
  geom_histogram(binwidth = 5, position = "dodge", color = "black") +
  labs(title = "Histogram of Aorta Diameter by Gender", x = "Diameter (mm)", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "lightblue"))

# Scatter plot of diameter v age - > colored by gender
ggplot(marfan, aes(x = age, y = diameter, color = sexe)) +
  geom_point(alpha = 0.6) +
  labs(title = "Aorta Diameter vs Age by Gender",
       x = "Age (years)", y = "Aorta Diameter (mm)", color = "Gender") +
  theme_minimal()

# Correlation between age and diameter
correlation <- cor(marfan$age, marfan$diameter)
cat("Correlation between age and diameter:", round(correlation, 3), "\n")
```

# 2. Fit Mixed-Effects Regression Model and Check the Assumptions
# ---------------------------------------------------------------

```{r}
# a linear mixed-effects model with random intercepts
model1 <- lmer(diameter ~ age + sexe + (1 | patnr), data = marfan, REML = FALSE)

# a linear mixed-effects model with random intercepts and slopes
model2 <- lmer(diameter ~ age + sexe + (age | patnr), data = marfan, REML = FALSE)
```
Comparing models using likelihood ratio test
```{r}
anova(model1, model2)
```
We'll proceed with model2
```{r}
summary(model2)
```
# Assumptions Checking
# ---------------------

Residuals vs Fitted Values
```{r}
# Extract residuals and fitted values
residuals_model2 <- residuals(model2)
fitted_model2 <- fitted(model2)

# residuals vs fitted values
ggplot(data.frame(fitted = fitted_model2, residuals = residuals_model2), aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs Fitted Values (Model 2)",
       x = "Fitted Values", y = "Residuals") +
  theme_minimal()
```
Normality of Residuals
```{r}
qqnorm(residuals_model2)
qqline(residuals_model2, col = "red")
```
Homoscedasticity
```{r}
# residuals v age
ggplot(data.frame(age = marfan$age, residuals = residuals_model2), aes(x = age, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs Age (Model 2)",
       x = "Age (years)", y = "Residuals") +
  theme_minimal()
```

# 3. Expected Aorta Diameter for New Patients Aged 20-50
# ------------------------------------------------------
Prepare New Data
```{r}
newdata <- expand.grid(age = seq(20, 50, by = 1),
                       sexe = factor(c("Female", "Male"), levels = levels(marfan$sexe)))

newdata$patnr <- NA
```
Predict Using Fixed Effects Only
```{r}
# Predicting diameter for new patients using only fixed effects.
newdata$pred_diameter <- predict(model2, newdata, re.form = NA)

X_new <- model.matrix(~ age + sexe, data = newdata)
fixed_effects <- fixef(model2)
vcov_matrix <- vcov(model2)

# SE
newdata$se_fit <- sqrt(diag(X_new %*% vcov_matrix %*% t(X_new)))

# CI
newdata$pred_lower <- newdata$pred_diameter - 1.96 * newdata$se_fit
newdata$pred_upper <- newdata$pred_diameter + 1.96 * newdata$se_fit
```
Plotting
```{r}
ggplot(newdata, aes(x = age, y = pred_diameter, color = sexe, fill = sexe)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = pred_lower, ymax = pred_upper), alpha = 0.2, color = NA) +
  labs(title = "Expected Aorta Diameter for New Patients",
       x = "Age (years)", y = "Aorta Diameter (mm)",
       color = "Gender", fill = "Gender") +
  theme_minimal()
```

# 4. Expected Aorta Diameter for Current Patients One Year Later
# --------------------------------------------------------------
Preparation
```{r}
last_measurements <- marfan %>%
  group_by(patnr) %>%
  filter(age == max(age)) %>%
  ungroup()

# The new data
last_measurements$age_plus_one <- last_measurements$age + 1

# Preparing data for prediction
predict_data <- last_measurements %>%
  select(patnr, sexe, age_plus_one) %>%
  rename(age = age_plus_one)
```
We will use bootstrapping to obtain prediction intervals for the aortic diameter one year later.
```{r}
# Function to predict for bootstrap samples
predict_boot <- function(fit) {
  predict(fit, newdata = predict_data, allow.new.levels = TRUE)
}

n_boot <- 1000

# Bootstrap predictions
set.seed(42)
boot_results <- bootMer(model2, 
                        FUN = predict_boot, 
                        nsim = n_boot, 
                        re.form = ~(age | patnr), 
                        use.u = TRUE, 
                        type = "parametric", 
                        parallel = "no")

pred_matrix <- t(boot_results$t)

# prediction intervals
predict_data$pred_diameter <- apply(pred_matrix, 1, mean)
predict_data$pred_lower <- apply(pred_matrix, 1, quantile, probs = 0.025)
predict_data$pred_upper <- apply(pred_matrix, 1, quantile, probs = 0.975)

```
Merge with original data
```{r}
plot_data <- last_measurements %>%
  select(patnr, age, sexe, diameter) %>%
  rename(original_age = age, original_diameter = diameter) %>%
  left_join(predict_data, by = c("patnr", "sexe"))
```
Plot
```{r}
ggplot(plot_data) +
  # I put lines connecting original and predicted points for clarification
  geom_segment(aes(x = original_age, y = original_diameter, 
                   xend = age, yend = pred_diameter, color = sexe, group = patnr),
               size = 0.5) +
  # orginal data points
  geom_point(aes(x = original_age, y = original_diameter, color = sexe), shape = 16, size = 2) +
  # one year later points
  geom_point(aes(x = age, y = pred_diameter, color = sexe), shape = 17, size = 2) +
  # error bars for one year later points
  geom_errorbar(aes(x = age, ymin = pred_lower, ymax = pred_upper, color = sexe), width = 0.2) +
  labs(title = "Original and Predicted Aorta Diameter One Year Later for Each Patient",
       x = "Age (years)", y = "Aorta Diameter (mm)", color = "Gender") +
  theme_minimal()
```

