---
title: 'week 3 assignment: aorta diameter growth in Marfan syndrome'
author: "Robin Manz, Matias Tolppanen"
date: "2024-11-12"
output: html_notebook
---
## Librarys 
```{r setup, include=FALSE}
# librarys
knitr::opts_chunk$set(echo = TRUE)
library(lme4)
library(merTools)
library(flexplot)
# install.packages("devtools")
# devtools::install_github("dustinfife/flexplot")
```
## Import 

```{r marfan}
data <- read.csv("marfan.csv", header=TRUE, stringsAsFactors=FALSE)
```

A1: Start by giving a data description.
## Descriptive
| Variable | Description | 
| --- | --- | 
| patnr | number of the patient (1, 2, 3,..., 159) | 
| metingnr | countvariable denoting the number of the diameter |
| **measurement in the same patient** |
| diameter | aortadiameter measurement (mm) |
| age | age at the time of the measurement (years) |
| sexe | gender of the patient: 1=male, 0=female | 

q1: what is the average growth rate per year
does the average growth vary with age
does the average growth rates differ between men and women
which patients show fast(er) growth and which patients show ‘stable’ disease
predict aorta diameter at various ages for new patients
predict next year’s aorta-diameter for the current patients



```{r pressure, echo=FALSE}
# Basic descriptives
n_pat <- length(unique(data$patnr))
n_pat_female <- length(unique(subset(data, sexe == 0)$patnr))
n_pat_male <- length(unique(subset(data, sexe == 1)$patnr))
print(paste("Number of entries:", nrow(data)))
print(paste("Number of unique patiens", n_pat))
hist(data$metingnr, main=paste("number of the measurement mean:", mean(data$metingnr)))
hist(data$diameter, main=paste("diameter of the aortas mean:", mean(data$diameter)))
hist(data$age, main=paste("Age mean:", mean(data$age) ))

# missing values
print(sum(is.na(data$column_name)))
# number of encounters per patient
n_pat_encounters <- seq(0,0,length.out=n_pat)
for (row in 1:nrow(data)) {
  patnr <- data[row, "patnr"]
  n_pat_encounters[patnr] <- n_pat_encounters[patnr] +1
}
hist(n_pat_encounters, main="encounters per patient")

# average growth rate per year. 
# they are sorted per patientnr
growths_per_year <- numeric()
last_diameter <- -1
last_patnr <- -1
last_age <- -1
for (row in 1:nrow(data)) {
  if (last_patnr == data[row, "patnr"]) {
    # case same patient as the last time
    dia_diff <- data[row, "diameter"] - last_diameter
    age_diff <- data[row, "age"] - last_age
    # diff per year 
    # Calculate difference per year
    if (age_diff != 0) {  # Avoid division by zero
      diff_per_year <- dia_diff / age_diff
      growths_per_year <- c(growths_per_year, diff_per_year)  # Append to the vector
    }
    
  } else {
    # case new patient
    # save initial data
    last_diameter <- data[row, "diameter"]
    last_patnr <- data[row, "patnr"]
    last_age <- data[row, "age"]
  }
}
print(paste("Average growth per year", mean(growths_per_year)))

```

## Check linearity of age and diameter

```{r}
library(lme4)
library(nlme)

# spagetti plot ?
plot(data$age, data$diameter, main = "Age vs Diameter")

linear_model <- glm(diameter ~ age, data = data)
summary(linear_model)
#m1 <- lme(diameter ~ age, data=data)

# fit a linear model
model <- lm(diameter ~ age + sexe, data = data)
# Plot residuals vs. fitted values
plot(fitted(model), resid(model), 
     xlab = "Fitted Values", 
     ylab = "Residuals", 
     main = "Residuals vs. Fitted Values")
abline(h = 0, col = "red")
```

Since there is no significant deviation with the increase of age we assume linearity 
-> lmer is to be used

## Random vs fixed effect explanation

```{r}

# Load necessary library
library(lme4)

# Example dataset (assume `data` exists with columns: diameter, age, sex, and group)
# `diameter`: Diameter of the aorta (response)
# `age`: Age of the individual (fixed effect)
# `sex`: Sex of the individual (fixed effect)
# `group`: Grouping variable (e.g., study center)

# Fixed Intercept and Fixed Slope (no random effects)
# Here, we assume the relationship between diameter, age, and sex is the same across all groups.
model_fixed <- lm(diameter ~ age + sex, data = data)

# Random Intercept Model
# Each group gets its own intercept, but the slopes (effects of age and sex) are the same for all groups.
model_random_intercept <- lmer(diameter ~ age + sex + (1 | group), data = data)

# Explanation:
# (1 | group): Adds a random intercept for each level of `group`.

# Random Intercept and Random Slope Model
# Each group gets its own intercept AND its own slope for `age`.
model_random_intercept_slope <- lmer(diameter ~ age + sex + (1 + age | group), data = data)

# Explanation:
# (1 + age | group): Adds a random intercept AND a random slope for `age` within each `group`.

# Random Slope Only (No Random Intercept)
# Each group gets its own slope for `age`, but the intercept is fixed across groups.
model_random_slope_only <- lmer(diameter ~ age + sex + (0 + age | group), data = data)

# Explanation:
# (0 + age | group): Adds only a random slope for `age` without a random intercept.

# Fixed Effects Only
# When random effects are unnecessary, you can use fixed effects alone.
model_fixed_effects_only <- lmer(diameter ~ age + sex, data = data)

```
## creating the best model
```{r} 
# baseline model, no predictor, just means
baseline_model <- lmer(diameter ~ 1 + (1|patnr), data = data )
# add age to the model, age is random effect
# random slope and random intercept
age_model <- lmer(diameter ~ age + (age|patnr), data = data)

print("comparison baseline model vs age model")
model.comparison(baseline_model, age_model)
# the age model is widely favourable

sex_model <- lmer(diameter ~ sexe + (sexe | patnr), data = data)
print("comparison baseline model vs sex_model")

model.comparison(baseline_model, sex_model)
# Sex model is better


#### Figure out if age should be fixed or not
# fixed age model same slope but different intercept
fixed_age_model <- lmer(diameter ~ age + (1| patnr), data = data)

visualize(age_model, plot="model", sample=50)
visualize(fixed_age_model, plot="model", sample=50)

print("comparison age_model vs fixex_age_model")
model.comparison(age_model, fixed_age_model)

## the age model is favoured by p-value in the comparison, AIC and bayes factor
estimates(age_model)

#### Figure out if sex should be fixed or not
fixed_sex_model <- lmer(diameter ~ sexe + (1| patnr), data= data)

visualize(sex_model, plot="model", sample=50)
visualize(fixed_sex_model, plot="model", sample=50)

print("comparison sex_model vs fixed_sex_model")
model.comparison(sex_model, fixed_sex_model)
# fixed model is not significantly better than the sex model


# get a plot for comparisons
compare.fits(diameter ~ age | patnr, data = data, model1 = baseline_model, model2= age_model, clusters = 10)

# add sex to model
#age_sex_model <- lmer(diameter ~ age + sexe + (1|patnr))
```

Results: 
- age should be modeled as random effect (age + (age|patnr))
- sexe should be modeled as fixed effect (sexe + (1|patnr)) also sex does not vary within clusters so it has to be fixed


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## see if splines improve the model
```{r}
library(splines)
aics <- c()
aicss <- c()

for (fixed_spline in 1:5) {
  for (random_spline in 1:5) {
    formula <- paste("diameter ~ ns(age, df=", fixed_spline, ") + sexe + ( ns(age, df=", random_spline , ")|patnr)")
    model <- lmer(formula, data = data)
    aic <- AIC(model)
    aics[[length(aics) + 1]] = c(aic, formula)
  }
}

print(aics)
min_tuple <- aics[[which.min(sapply(aics, function(x) x[1]))]]

# bes model according to this: 
# "4822.16353576645"    "diameter ~ ns(age, df= 4 ) + sexe + ( ns(age, df= 5 )|patnr)"


best_splines_model <- lmer(diameter ~ ns(age, df= 4 ) + sexe + ( ns(age, df= 5 )|patnr), data = data)
plot(fitted(best_splines_model), resid(best_splines_model), 
     xlab = "Fitted Values", 
     ylab = "Residuals", 
     main = "Residuals vs. Fitted Values")
abline(h = 0, col = "red")


# m0_ns <- lmer(diameter ~ ns(age, df=4) * sexe + (1+age|patnr), data=dat)

AIC(best_splines_model)

age_seq <- seq(min(data$age), max(data$age), by=0.1)  # adjust step size for smoother lines
new_data <- expand.grid(age = age_seq, sexe = c(0, 1))

new_data$pred_diameter <- predict(best_splines_model, newdata=new_data, re.form=NA)

library(ggplot2)
ggplot(data, aes(x=age, y=diameter)) +
  # original data points in gray
  geom_point(colour = "gray") +
  
  geom_line(data=new_data, aes(x=age, y=pred_diameter, colour=as.factor(sexe)), size=1.2) +
  scale_colour_manual(values = c("0" = "salmon", "1" = "palegreen3"), 
                      labels = c("Female", "Male"), 
                      name = "Sex") +
  labs(x = "Age", y = "Diameter", title = "diameter ~ ns(age, df= 4 ) + sexe + ( ns(age, df= 5 )|patnr)") #+
#  theme_minimal() +
#  theme(legend.position = "right")
```


## Model with both variables
```{r}

full_model <- lmer(diameter ~ age + sexe + (age|patnr), data = data)

model.comparison(full_model, age_model)

# Full model is significantly better based on AIC and bayes factor

compare.fits(diameter~ age | sexe + patnr, data=data, model1=age_model, model2=full_model)

summary(full_model)
# the average diameter for someone of average age and sex
# 27.18633 every year on average increases the diameter by 0.39761
# sex == 1 (male) increases the average diameter by 8.46241
estimates(full_model)
# plot residuals for linearity
plot(residuals(full_model))
plot(residuals(age_model) ~ data$age)

# include an interaction between sex and age and see if it is a better model than without
age_sex_inter_model <- lmer(diameter ~ age*sexe + (age|patnr), data = data)

model.comparison(full_model, age_sex_inter_model)
# AIC, bic and bayes factor favor the full model by a tiny bit, but the comparison is not significant. 
# This indicates that interactions between age and sex are not significant, hence there is no different growth rates between women and men

```

## Answers

- the average growth per year of age is 0.39761
- the average growth rate does not vary with age since the relationship between age and diameter is linear. Plotting the residuals against age it can be seen that they are not increasing in time, indicating a adequate fit of the linear model. 
- the average growth rate is not significantly different between men and women. The model with the interaction term does not perform significantly better than the model without the interaction


## get slopes per patient
```{r}

# Extract the random effects for age (growth rate) for each patient
random_effects <- ranef(full_model)$patnr

# Calculate the patient-specific slopes by adding the fixed effect of age
# to each patient's random effect for age
fixed_slope <- fixef(full_model)["age"]
patient_slopes <- fixed_slope + random_effects$age

sorted_patient_slopes <- sort(patient_slopes, decreasing = TRUE)

top_patients <- names(sorted_patient_slopes)[1:5]

print(top_patients)
print(sorted_patient_slopes[1:5])


# Extract random effects for each patient and convert to a data frame
random_effects <- as.data.frame(ranef(full_model)$patnr)

# Add the fixed effect for age to each patient's random effect to get the patient-specific slopes
fixed_slope <- fixef(full_model)["age"]
random_effects$patient_slope <- fixed_slope + random_effects$age

# Now add the patient IDs as a column if not already included
random_effects$patnr <- rownames(random_effects)

# Sort by patient-specific slope to get the patients with the biggest increases in diameter
top_patients <- random_effects[order(-random_effects$patient_slope), ]

# View the top 5 patients with the largest growth rates
top_5_patients <- head(top_patients, 5)
top_5_patients

# View the last 5 pateitns with the slowerst growth rates
bottom_5_patients <- tail(top_patients, 5)
bottom_5_patients
```

Answers: 
- the top five patiens with diameter increase per year are: 108 (1.584), 76(1.298), 157(1.247), 55(1.240) and 77(1.230)
- the bottom five patients with diameter decrease per year are: 151(-0.28, only one measurement!), 123(-0.089), 95 (-0.034), 35(-0.010) and 2(0.01)

## Predictions for new patients

```{r}

set.seed(42)

# TODO plot that bitch
# predictions for new entries of existing data
sampled_data <- data[sample(nrow(data), 10), c("patnr", "age", "sexe")]

sampled_data$age <- sampled_data$age + 1

# Get predictions with confidence intervals for existing patients (including random effects)
updated_patient_predictions <- predictInterval(
  merMod = full_model,
  newdata = sampled_data,
  level = 0.95,               # Confidence level
  n.sims = 1000,              # Number of simulations
  which = "full"              # Full model including random effects
)
print(updated_patient_predictions)

sampled_data$predicted <- updated_patient_predictions$fit        # Add predicted values
sampled_data$lower_bound <- updated_patient_predictions$lwr     # Add lower bound of CI
sampled_data$upper_bound <- updated_patient_predictions$upr

# 5. Combine the original data and the modified data (new predictions)
# Assuming the original data has 'patnr', 'age', and 'sexe' columns
original_data <- data[, c("patnr", "age", "sexe")]
original_data$predicted <- NA      # No predictions for the original data
original_data$lower_bound <- NA    # No CI for the original data
original_data$upper_bound <- NA    # No CI for the original data

# Combine the original data with the modified sampled data
combined_data <- rbind(original_data, sampled_data)

# 6. Plot the original data and new predictions with confidence intervals
# 6. Plot the original data and new predictions with confidence intervals
ggplot(sampled_data, aes(x = age, group = patnr)) +
  # Plot the new predictions for each patient as separate lines
  geom_line(aes(y = predicted, color = as.factor(patnr)), size = 1) +
  
  # Add the confidence interval ribbon for each patient's prediction
  geom_ribbon(aes(ymin = lower_bound, ymax = upper_bound, fill = as.factor(patnr)), 
              alpha = 0.2) +
  
  # Customize the plot
  labs(x = "Age", y = "Predicted Diameter", 
       title = "New Predictions with 95% Confidence Intervals for Each Patient") +
  scale_color_manual(values = rainbow(length(unique(sampled_data$patnr)))) +
  scale_fill_manual(values = rainbow(length(unique(sampled_data$patnr)))) +
  theme_minimal() +
  theme(legend.title = element_blank())


# Create new data frame for new patients without patnr (new patients)
new_data <- data.frame(
  age = seq(20, 80, by = 5),   # Sequence of ages
  sexe = rep(1, 13),
  patnr = rep(250, 13)# Assume all are male
)

# Get predictions with confidence intervals for new patients (fixed effects only)
new_patient_predictions <- predictInterval(
  merMod = full_model,
  newdata = new_data,
  level = 0.95,              # Confidence level for interval
  n.sims = 1000,             # Number of simulations for interval estimation
  which = "fixed"            # Fixed effects only, no random effects
)

print(new_patient_predictions)


### Plotting

library(ggplot2)
# Create a sequence for the x-axis (e.g., age or any variable that corresponds to the rows)
x_axis <- seq(20, 80, by = 5)  # Adjust as necessary

# Add the x_axis to the result_df
new_patient_predictions$x_axis <- x_axis

# Plot the predicted values and the confidence intervals
ggplot(new_patient_predictions, aes(x = x_axis)) +
  geom_line(aes(y = fit), color = "blue", size = 1) +  # Predicted values
  geom_ribbon(aes(ymin = upr, ymax = lwr), fill = "gray", alpha = 0.5) +  # Confidence interval
  labs(x = "Age", y = "Predicted Diameter") +
  theme_minimal() +
  ggtitle("Predicted Values with 95% Confidence Interval")
```